---
title: "Occurrence data from PaleoBioDB"
---

In the previous topic, we learned that it is possible to plot points on a map providing their coordinates as arguments in the `points()` function. However, plotting point-by-point would be meaningless when dealing with larger datasets and compiling their coordinates (or paleocoordinates) can be very time-consuming. Luckily, such data has, in many cases, been compiled by other researchers and, sometimes, made available in larger online repositories. Examples of those repositories are the Global Biodiversity Information Facility or [GBIF](https://www.gbif.org/) for data on biodiversity and the Paleobiology Database or [PaleoBioDB](https://paleobiodb.org/#/), specifically for occurrences of fossils. In this page, we will focus on the latter, exploring the R package `paleobioDB` to learn how to download and work with occurrence data of fossils using R.

```{r}
#| results: hide
#| message: false
#| echo: false
library(paleobioDB)
library(palaeoverse)
library(viridisLite)
library(terra)
```

In order to download occurrence data from PaleoBioDB, we will use the function `pbdb_occurrences()`. This function allows us to specify various parameters to filter the data we want to download, such as taxonomic group, time interval, geographic location, and more, and also to select which data to download. Let's download a dataset composed of Cenozoic *Testudines* (turtles) occurrences:

```{r}
#| results: hide
testudines <- pbdb_occurrences(base_name = "Testudines",
                               interval = "Cenozoic",
                               show = c("coords", "classext", "paleoloc"),
                               limit = "all", 
                               vocab = "pbdb")  ## vocab makes better names

```

```{r}
dim(testudines)
```
This dataset is composed of 5911 occurrences and it has 35 columns, with different types of information about each occurrence. The `paleobioDB` package includes a function for mapping the occurrences directly, called `pbdb_map()`. This function includes arguments for changing the mainland and ocean colors (`col_int` and `col_ocean`, respectively), and another (`col_point`) for changing the color gradient for the occurrences. Let's explore this function and its arguments to plot the occurrences we just downloaded:
```{r}
# basic plot
pbdb_map(data = testudines,
         main = "Cenozoic Testudines occurrences")

# changing colors
pbdb_map(data = testudines,
         main = "Cenozoic Testudines occurrences",
         col_int = "grey80",
         col_ocean = "lightblue",
         col_point = c("lightyellow", "yellow", "orange", "red"),
         cex = 0.8)

# using the color pallette from the package "viridisLite"
pbdb_map(data = testudines,
         main = "Cenozoic Testudines occurrences",
         col_int = "grey80",
         col_ocean = "lightblue",
         col_point = viridis(5))
```

Instead of plotting single points for each occurrence, it is also possible to create richness maps, which can be useful to visualize areas with higher or lower taxonomic richness. To do so, we can use the function `pbdb_map_richness`, in which we need to select which taxonomic rank (e.g., species, genus, family, etc.) we want to consider. Let's see how this works using different taxonomic ranks and color palettes:
```{r}
#| message: false
#| warning: false
# by species
map.rich <- pbdb_map_richness(data = testudines, 
                              rank = "species",
                              res = 5,
                              col_int = "grey80",
                              col_ocean = "lightblue",
                              col_rich = heat.colors(10),
                              title = "Species richness")

# by genera
map.rich <- pbdb_map_richness(data = testudines, 
                              rank = "genus",
                              res = 5,
                              col_int = "grey80",
                              col_ocean = "lightblue",
                              col_rich = viridis(10),
                              title = "Genus richness")

# by family
map.rich <- pbdb_map_richness(data = testudines, 
                              rank = "family",
                              res = 5,
                              col_int = "grey80",
                              col_ocean = "lightblue",
                              col_rich = map.pal("inferno",10),
                              title = "Family richness")
```

This page provided a brief overview of how to download and plot occurrence data from the Paleobiology Database using R. The `paleobioDB` package offers many more functions and options for working with paleontological data, so I encourage you to explore its documentation for further details and possibilities. Let's take a look at some additional resources from the package.
```{r}
#| message: false
#| warning: false
# Temporal range of taxa
temp.range <- pbdb_temp_range(testudines,
                              rank = "family",
                              col = "#AA55AA") ## select color

# Richness through time
rich <- pbdb_richness(testudines, 
                      rank = "species", 
                      ylab = "Number of Species", 
                      temporal_extent = c(0, 23)) ## define temporal extent

# change color the resolution, default = 1
rich <- pbdb_richness(testudines, 
                      rank = "genus", ## change rank
                      ylab = "Number of Species", 
                      temporal_extent = c(0, 23),
                      res = 0.5, ## looks more detailed
                      col = "#A0A00F30",
                      bord = "#A0A00F")  


# Patterns of origination and extinction
origination <- pbdb_orig_ext(testudines,
                      rank = "genus",
                      temporal_extent = c(0, 66),
                      orig_ext = 1,  ## this argument defines whether origination
                      res = 2)

extinction <- pbdb_orig_ext(testudines,
                            rank = "genus",
                            temporal_extent = c(0, 66),
                            orig_ext = 2,  ## or extinction
                            res = 2)
```

We can further explore the dataset downloaded from the PaleoBioDB with the package `palaeoverse`. [Palaeoverse](https://palaeoverse.org/) is an community initiative to share open-source tools for paleontological data analysis and visualization in R. It includes different packages, but here we are exploring only `palaeoverse`, which provides functions to prepare and explore paleontological data. Let's see how we can use some of its functions to explore the `testudines` dataset:
```{r}
cenozoic.bins <- time_bins(interval = "Cenozoic",
                           plot = TRUE)

```

The package includes another useful function called `bin_time` which can be used to assign occurrences to time bins. This can be particularly useful for analyzing temporal patterns in the data using a common time framework. With this information, we can then run the function `tax_range_time()` to plot temporal ranges using our dataset.
```{r}
family_range <- tax_range_time(subset(testudines, !is.na(family)),
                               name = "family",
                               plot = TRUE,
                               plot_args = list(ylab = "Families",
                                                pch = 15,
                                                col = "grey30",
                                                lty = 2))
```

`palaeorotate()` is a function included in the `palaeoverse` package that allows us to calculate paleocoordinates for fossil occurrences based on their present-day coordinates and age. This can be particularly useful for ploting past distributions on paleogeographic maps. Let's see how we can use this function with our `testudines` dataset, but before we proceed we need to assign the occurrences to a common framework of time bins, which can be done using the function `bin_time()`. Then we can use the `look_up()` function to convert the interval values derived from the PaleoBioDB into official numeric ages based on the intervals of the International Comission on Stratigraphy (ICS). This is useful for standardizing data from different sources. After this we can finally plot those occurrences per time bin and per paleocoordinates.
```{r}
#| warning: false
# assing occurrences to time bins
testudines.binned <- bin_time(occdf = testudines, 
                              bins = cenozoic.bins,
                              method = "majority")
# calculate paleocoordinates
testudines.palaeocoord <- palaeorotate(testudines.binned, 
                                       age = "bin_midpoint")  ## using the midpoint of their range

# convert to ICS intervals
testudines.palaeocoord <- look_up(testudines.palaeocoord, 
                                  early_interval = "early_interval", 
                                  late_interval = "late_interval", 
                                  int_key = interval_key)

# plot occurrences per time bin and per paleocoordinates
plot(x = testudines.palaeocoord$interval_mid_ma,
     y = testudines.palaeocoord$p_lat,
     xlab = "Time (Ma)",
     ylab = "Palaeolatitude (\u00B0)",
     xlim = c(66, 0), 
     xaxt = "n",
     pch = 20,
     cex = 1.5)

# and we can add a geological time scale to the plot
axis_geo(side = 1, 
         intervals = "epoch")
```

