---
title: "Saving plots"
---

Once you have created a plot in R, you may want to save it for future use or to include it in a report or presentation. R provides several functions to save plots in different formats, such as PNG, JPEG, PDF, and SVG. Let's create a simple plot and save it as PNG (image) and PDF (vectorized) formats.

```{r}
plot(1:10, 1:10, main = "My Plot", xlab
     = "X-axis", ylab = "Y-axis")
```

To save a plot as a PNG file, you can use the `png()` function to specify the file name and dimensions, followed by the plotting commands, and then use the `dev.off()` function to close the device and save the file. Here's an example:
```{r}
#| output: false
# Create a PNG file
png("my_plot.png", width = 800, height = 600)
# Create a simple plot
plot(1:10, 1:10, main = "My Plot", xlab
     = "X-axis", ylab = "Y-axis")
# Close the device
dev.off()
```

To save a plot as a PDF file, you can use the `pdf()` function in a similar way:
```{r}
#| output: false
pdf("my_plot.pdf", width = 8, height = 6)

plot(1:10, 1:10, main = "My Plot", xlab
     = "X-axis", ylab = "Y-axis")
# Always close the device
dev.off()
```

These commands will create two files, `my_plot.png` and `my_plot.pdf`, in your working directory. You can adjust the width and height parameters to change the size of the saved plots as needed.
